---
output: 
  pdf_document:
    fig_caption: true
title: "Project Report: A Reproduction Analysis of Allele Frequency Differences Between Populations"
thanks: "The original study, **Population differentiation in allele frequencies of obesity-associated SNPs**, is available in [BMC Genomics] (https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-017-4262-9#Equ1)."
author:
- name: Bianca Kirsh
  affiliation: University of British Columbia
- name: Josh de Freitas
  affiliation: University of British Columbia
abstract: "This document presents the re-analysis of a published genomics study, including the written report, integrated code, and graphs."
keywords: "genomics, analysis"
date: "`r format(Sys.time(), '%B %d, %Y')`"
geometry: margin=1in
fontfamily: mathpazo
fontsize: 11pt
# spacing: double
endnote: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      echo=TRUE,
                      message=FALSE,
                      warning=FALSE,
                      fig.path='figs/',
                      cache.path = '_cache/',
                      fig.process = function(x) {
                      x2 = sub('-\\d+([.][a-z]+)$', '\\1', x)
                      if (file.rename(x, x2)) x2 else x
                      })
```

```{r}
# install any of these that you might not have already
library(readxl)
library(reshape2)
library(ggplot2)
library(dplyr)
library(tidyr)
library(ggvenn)
library(heatmap3)
```

# Introduction

## Overview of the original experiment

Large research efforts have been made to identify genomic regions, variants in candidate genes, and environmental factors that contribute to disease-associated health disparities across populations.  

In a study by [Mao et al.](https://bmcgenomics.biomedcentral.com/articles/10.1186/s12864-017-4262-9#Equ1), worldwide differences in the effect allele frequency of 225 obesity-associated SNPs were investigated. The original analysis was conducted to identify significantly enriched or depleted effect alleles across 26 populations, which were later interrogated with respect to the relevance and underlying biological mechanisms of their associated genes. Additionally, researchers inspected the population-based patterns that emerged from differences in allele frequency and genetic risk scores. The results suggest that over 85% of effect alleles exhibit significant allele frequency differences and that population-level differences in genetic risk scores are correlated with obesity prevalence.  

## Goals of the re-analysis

In Genome-Wide Association Studies (GWAS), the over-representation of European populations is frequently coupled with inadequate sampling of African genomes in SNP arrays and imputation reference panels. As a result, ascertainment bias commonly confounds the associations between genetic ancestry and disease susceptibility.  

The main goal of this re-analysis is to explore obesity-related variants among a subset of samples, namely those associated with the African American (ASW), African (YRI), and European American (CEU) populations. As such, this project will interrogate the relationship between disease susceptibility, ancestry, and admixture to elucidate the genetic underpinnings of health disparities across populations. Furthermore, this re-analysis will evaluate the transferability of GWA studies by considering the ancestral diversity within and between study cohorts.  

Ultimately, this re-analysis will illustrate the spectrum of genetic diversity between geographically distant genomes to reveal ancestral biological signatures and systematic differences underlying the obesity epidemic. In doing so, we hope to encourage diversity and inclusion in genomics research, as well as urge for the investigation of genetic variants with extreme allele frequency differences (EAFD) between populations.  

# Methods

For this project, we first obtained a list of obesity-associated SNPs and their effect allele frequencies from the supplementary material available in the original publication.  The authors compiled 225 SNPs from 29 GWA studies in the NHGRI-EBI GWAS Catalog, 19 of which were performed in European populations, 3 in East Asians, 2 in South Asians, 3 in Africans, and 2 in mixed ethnic populations. The corresponding effect allele frequencies were based on genotype information for 26 populations surveyed by the 1000 Genomes Project.

```{bash, eval = TRUE}

#
wget https://static-content.springer.com/esm/art%3A10.1186%2Fs12864-017-4262-9/MediaObjects/12864_2017_4262_MOESM1_ESM.xlsx

```

The remaining steps implement a pipeline that uses the R programming language to automate key components of data processing, analysis, and visualisation. 

## Overview of the pipeline stages


We then selected a subset of the original populations to compare the association signals of each variant between African American (ASW), African (YRI), and European American (CEU) populations. The re-analysis consisted of 3 main interrogations, namely investigating significant enrichment or depletion patterns, extreme allele frequency differences, and the biological mechanisms of relevant genes as they relate to admixture. 


```{r, eval = TRUE, tidy = TRUE, cache=FALSE, message=F, warning=F}



```


## Quality Control


## Analysis


```{r}

```

# Results

# Conclusions